---
title: "part I - Keras, Tensorflow and MLP"
author: "Michał Maj"
output: html_notebook
---

Before we start building neural networks using Keras let's see how R communicates with python.

```{r python_config}
library(keras)
library(reticulate)
library(tidyverse)
# install_keras() Keras installation form R
py_config() # Current config info
# Using commands below you can set up correct python path/env:
# use_python("/usr/local/bin/python")
# use_virtualenv("~/myenv")
# use_condaenv("myenv")
# You can also use RETICULATE_PYTHON system variable
Sys.getenv("RETICULATE_PYTHON")
readLines(".Rprofile")
```

```{r reticulate_example}
np <- import("numpy")
np$max(c(4, 7, 2))
```

In Keras we can create models in two different ways:
 - build sequential model - we're stacking new layers on top of previous ones. We can't use multiple inputs and outputs.
  - using functional API - allows us to use multiple inputs and outputs.

We will start with sequential model. We have to start with a model initialization:

```{r sequential_model}
load("data/boston.RData")
# Check shape of the data
boston_train_X %>% dim() # Two dim tensor
boston_train_Y %>% dim() # One dim tensor

boston_model <- keras_model_sequential()
```

In the next step we can add some layers (note that we don't have to reassign the model with `<-`):

```{r add_layer}
boston_model %>% layer_dense(units = 16, # Number of neurons in the layer
                      activation = "tanh", # Activation function
                      input_shape = c(13)) # Nr of predictors. always in first layer!
boston_model
```

```{r nr_of_params}
# Explain why do we have 224 params ?
13 * 16 + 16
```

After adding hidden layer we can add output layer:

```{r}
boston_model %>%
  layer_dense(units = 1,
              activation = "linear")
boston_model
```

We can now configure model for training. We will use SGD as optimizer, MSE as loss function and add MAE as additional metric.

```{r model_compilation}
boston_model %>% compile(
  optimizer = "sgd",
  loss = "mse",
  metrics = c("mae")
)
```

We are ready to train our first neural network:

```{r model_training}
history <- boston_model %>%
  fit(x = boston_train_X,
      y = boston_train_Y,
      validation_split = 0.2, # 20% of the data for validation
      epochs = 5, # Number of "loops" over whole dataset
      batch_size = 30, # Sample size for one run of SGD
      verbose = 1)
```

We can now evaluate trained model on the test dataset:

```{r model_evaluation}
boston_model %>%
  evaluate(boston_test_X, boston_test_Y)
```

And calculate predictions:

```{r prediction}
boston_predictions <- boston_model %>% predict(boston_test_X)
head(boston_predictions)
```

In the end we can save our model on hard drive:

```{r save_model}
save_model_hdf5(boston_model, "boston_model.hdf5")
```


