---
title: "part I - Keras, Tensorflow and MLP"
author: "Michał Maj"
output: html_notebook
---

Before we start building neural networks using Keras let's see how R communicates with python.

```{r python_config}
library(keras)
library(reticulate)
library(tidyverse)
library(gridExtra)
# install_keras() Keras installation form R
py_config() # Current config info
# Using commands below you can set up correct python path/env:
# use_python("/usr/local/bin/python")
# use_virtualenv("~/myenv")
# use_condaenv("myenv")
# You can also use RETICULATE_PYTHON system variable
Sys.getenv("RETICULATE_PYTHON")
readLines(".Rprofile")
```

```{r reticulate_example}
np <- import("numpy")
np$max(c(4, 7, 2))
```

In Keras we can create models in two different ways:
 - build sequential model - we're stacking new layers on top of previous ones. We can't use multiple inputs and outputs.
  - using functional API - allows us to use multiple inputs and outputs.

We will start with sequential model. We have to start with a model initialization:

```{r sequential_model}
load("data/boston.RData")
# Check shape of the data
boston_train_X %>% dim() # Two dim tensor
boston_train_Y %>% dim() # One dim tensor

boston_model <- keras_model_sequential()
```

In the next step we can add some layers (note that we don't have to reassign the model with `<-`):

```{r add_layer}
boston_model %>% layer_dense(units = 16, # Number of neurons in the layer
                      activation = "tanh", # Activation function
                      input_shape = c(13)) # Nr of predictors. always in first layer!
boston_model
```

Why do we have 224 params ?

```{r nr_of_params}
13 * 16 + 16
```

After adding hidden layer we can add output layer:

```{r add_output_layer}
boston_model %>%
  layer_dense(units = 1,
              activation = "linear")
boston_model
```

We can now configure model for training. We will use SGD as optimizer, MSE as loss function and add MAE as additional metric.

```{r model_compilation}
boston_model %>% compile(
  optimizer = "sgd",
  loss = "mse",
  metrics = c("mae")
)
```

We are ready to train our first neural network:

```{r model_training}
history <- boston_model %>%
  fit(x = boston_train_X,
      y = boston_train_Y,
      validation_split = 0.2, # 20% of the data for validation
      epochs = 5, # Number of "loops" over whole dataset
      batch_size = 30, # Sample size for one run of SGD
      verbose = 1)
```

We can now evaluate trained model on the test dataset:

```{r model_evaluation}
boston_model %>%
  evaluate(boston_test_X, boston_test_Y)
```

And calculate predictions:

```{r prediction}
boston_predictions <- boston_model %>% predict(boston_test_X)
head(boston_predictions)
```

In the end we can save our model on hard drive:

```{r save_model}
save_model_hdf5(boston_model, "boston_model.hdf5")
```

We know now how to use MLP for regression tasks, let's check how to use it for classification problems. We will start with binary classification:

```{r load_bin_class_data}
load("data/bin_class.RData")
ggplot(bin_class_data, aes(x, y, color = factor(class))) + theme_bw() + geom_point()
```

We have to transform the data into tensors:

```{r data_to_tensors}
ind <- sample(1:nrow(bin_class_data), 0.8*nrow(bin_class_data))
bin_class_train_X <- bin_class_data[ind, c("x", "y")] %>% as.matrix()
bin_class_train_Y <- bin_class_data[ind, "class", drop = TRUE]
bin_class_test_X <- bin_class_data[-ind, c("x", "y")] %>% as.matrix()
bin_class_test_Y <- bin_class_data[-ind, "class", drop = TRUE]
```

Let's create a few simple models:

```{r bin_model}
# TASK: Create a sequential model with:
# a) one hidden layer b) two hidden layers
# Check variants with 2, 4 and 8 units in first layer
# Check variants with relu, tanh, sigmoid activations
# Add l1/l2 regularization using 'activity_regularizer' argument and regularizer_l1/2() functions
# Compile model with binary crossentropy as loss and SGD optimizer
# Fit the model using 10% of the data for validation, 100 epochs and batch_size of 100. Save runs to 'history' object. Use 'verbose' = 0.
# Evaluate model on the test data

history$metrics$epoch <- 1:100 # Epochs
history_df <- as.data.frame(history$metrics)
ggplot(history_df, aes(epoch, acc)) + theme_bw() + geom_line(color = "red") + geom_line(aes(y = val_acc), color = "blue")
```

```{r bin_model_plot}
predictions <- bin_model %>% predict(square_data) %>% cbind(square_data) %>%
  as.data.frame() %>% set_names(c("proba", "x", "y"))
ggplot(predictions, aes(x, y)) + theme_bw() + geom_raster(aes(fill = proba)) + geom_contour(colour = "white", aes(z = proba)) + scale_fill_gradient(low = "red", high = "blue") + geom_point(data = bin_class_data, aes(x, y, color = factor(class)))
```

In a similar way we can build sequential model for multi-class classification problem:

```{r fashion_mnist}
load("data/fashion_mnist.RData")
xy_axis <- data.frame(x = expand.grid(1:28, 28:1)[, 1],
                      y = expand.grid(1:28, 28:1)[, 2])
plot_theme <- list(
  raster = geom_raster(hjust = 0, vjust = 0),
  gradient_fill = scale_fill_gradient(low = "white", high = "black", guide = FALSE),
  theme = theme_void()
)

sample_plots <- sample(1:nrow(fashion_mnist_train_X), 100) %>% map(~ {
  plot_data <- cbind(xy_axis, fill = data.frame(fill = fashion_mnist_train_X[.x, ]))
  ggplot(plot_data, aes(x, y, fill = fill)) + plot_theme
})

do.call("grid.arrange", c(sample_plots, ncol = 10, nrow = 10))
```

Beside l1/l2 regularization we can also use dropout. In this task you will build multi layer MLP with dropout regularization:

```{r fashion_mnist_model}
# TASK: Create MLP for fashion MNIST classification.
# Change labels vectors to one-hot-encoding matrix using to_categorical() function

# Scale pixel values to [0, 1] interval

# Model architecture:
# Dense layer with 512 units and "relu" activation
# Dropout layer with 20% drop rate
# Dense layer with 512 units and "relu" activation
# Dropout layer with 20% drop rate
# Output dense layer (how many units and what activation should You use?)

# Set SGD as optimizer and use categorical crossentropy as loss function. Use accuracy as additional metric.

# Fit the model. Use 20% of the data for validation, 20 epochs and 128 samples for batch size.

# Evaluate model on test set.
```

