---
title: "part VII - Recurrent Neural Networks"
author: "Michał Maj"
output: html_notebook
---

```{r packages}
library(keras)
library(tidyverse)
```

**Recurrent Neural Networks** can be used for many different tasks like language to language translation, time series forecasting, speach to text and text to speach translation, text/music generation, sequence classification/regression problems. We will start with a simple sentiment analysis (classification). We will use twitter data:

```{r sentiment140}
load("data/sentiment140.RData")
sentiment140_X %>% head()
```

First thing we have to do is to change regular text into tokens. We will start by creating tokenizer:

```{r sentiment140_tokenizer}
tokenizer <- text_tokenizer(
  num_words = 20000, # Max number of unique words to keep
  filters = "!\"#$%&()*+,-./:;<=>?@[\\]^_`{|}~\t\n", # Signs to filter out from text
  lower = TRUE, # Schould everything be converted to lowercse
  split = " ", # Token splitting character
  char_level = FALSE, # Should each sign be a token
  oov_token = NULL # Token to replace out-of-vocabulary words
)
sentiment140_X <- iconv(sentiment140_X, to = "UTF-8")
tokenizer %>% fit_text_tokenizer(sentiment140_X)
```

Now using tokenizer we can turn raw sentences into tokens:

```{r sentiment140_tokens}
sequences <- texts_to_sequences(tokenizer, sentiment140_X)
```

As you probably remember in Keras each sample has to be saved as tensor with the same shape. In case of RNN we are inserting sequences and each sequence has to be the same length. We have to pad our sequences:

```{r sentiment140_pad}
maxlen <- 50
sequences_pad <- pad_sequences(sequences, maxlen = maxlen)
```

Now we can split sequences into train and test set:

```{r sentiment140_split}
train_factor <- sample(1:nrow(sequences_pad), nrow(sequences_pad) * 0.8)
sentiment140_X_train <- sequences_pad[train_factor, ]
sentiment140_X_test <- sequences_pad[-train_factor, ]
sentiment140_Y_train <- sentiment140_Y[train_factor]
sentiment140_Y_test <- sentiment140_Y[-train_factor]
```

Let's think a little bit about words representation. In our case each word can be represented as a value between 0 and 20000, this value represents a key in the dictionary. For the numerical calculations it won't be enough... each word will be represented as on-hot encoded vector with the dimmensionality of 20000. That's a large number. There is a better way to represent words (and not only), we can use so called **embedings**. In Keras we can easly add to our neural network an embedding layer:

```{r sentiment140_embedding}
model <- keras_model_sequential() %>%
  layer_embedding(input_dim = 20000,
                  output_dim = 128, # Represent each word in 128-dim space
                  input_length = maxlen)
```

After embedding layer we can add recurrent layer an the output layer:

```{r sentiment140_rnn}
model %>%
  layer_simple_rnn(units = 32) %>%
  layer_dense(units = 1, activation = "sigmoid")
```

If the architecture is finished we can compile the model:

```{r sentiment140_compile}
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)
```

And fit it:

```{r sentiment140_fit}
history <- model %>% fit(
  sentiment140_X_train,
  sentiment140_Y_train,
  epochs = 10,
  batch_size = 128,
  validation_split = 0.2
)
```

We can see the overfitting, let's now try LSTM or GRU units instead:

```{r sentiment140_lstm}
# Ex. Expand exsisting model by changing simple RNN units for stacked LSTM (or GRU) layers.
# 1. Model architecture:
# Use same embedding layer
# Add LSTM layer with 15 units, recurrent dropout with 0.5 rate and don't forget to return sequences to LSTM on top
# Add LSTM layer with 7 units, recurrent dropout with 0.5 rate
# Add dense layer as output with 'sigmoid' activation
model2 <- ___

# 2, Comile the model

# 3. Fit the model

```

Now it's your turn, use the stack overflow dataset to classify each question to a programming language:

```{r so_questions}
load("data/stack_overflow.RData")

# Ex. Using knowlege from previous chapters create model for stack overflow question tags classification.
```

