---
title: "part VII - Recurrent Neural Networks"
author: "Michał Maj"
output: html_notebook
---

```{r packages}
library(keras)
library(tidyverse)
```

**Recurrent Neural Networks** can be used for many different tasks like language to language translation, time series forecasting, speach to text and text to speach translation, text/music generation, sequence classification/regression problems. We will start with a simple sentiment analysis (classification). We will use twitter data:

```{r sentiment140}
load("data/sentiment140.RData")
sentiment140_X %>% head()
```

First thing we have to do is to change regular text into tokens. We will start by creating tokenizer:

```{r sentiment140_tokenizer}
tokenizer <- text_tokenizer(
  num_words = 20000, # Max number of unique words to keep
  filters = "!\"#$%&()*+,-./:;<=>?@[\\]^_`{|}~\t\n", # Signs to filter out from text
  lower = TRUE, # Schould everything be converted to lowercse
  split = " ", # Token splitting character
  char_level = FALSE, # Should each sign be a token
  oov_token = NULL # Token to replace out-of-vocabulary words
)
tokenizer %>% fit_text_tokenizer(sentiment140_X)
```

